---
title: "H1-Exercise 1"
author: "Constanza F. Schibber"
date: "1/22/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Excercise 1 (11.4)

 The folder cd4 has CD4 percentages for a set of young children with HIV who were measured several times over a period of two years. The dataset also includes the ages of the children at each measurement.

```{r data}
hiv.data<-read.csv("http://www.stat.columbia.edu/~gelman/arm/examples/cd4/allvar.csv")
colnames(hiv.data)
hiv.data <- subset(hiv.data, (CD4PCT>0) & (baseage>0))
hiv.data$sqrt.CD4 <- sqrt(hiv.data$CD4PCT) 
hiv.data$time <- hiv.data$visage - hiv.data$baseage # create time

#create id variable to represent unique new id for each child
hiv.data$id<-NA
J<-length(hiv.data$newpid)
uniq.pid<-unique(hiv.data$newpid)
for (i in 1:J){
hiv.data$id[hiv.data$newpid==uniq.pid[i]] <- i
}

summary(hiv.data$id)
```

(a) Graph the outcome (the CD4 percentage, on the square root scale) for each child as a function of time. Select 4 children because there are many of them.

```{r 1.a}

```

(b) Each child’s data has a time course that can be summarized by a linear fit. Estimate these lines and plot them for all the children. (Do it for the 4 children you selected above)

```{r 1.b unpooled model}

```

# Exercise 2 (12.2)

(a) Write a model predicting CD4 percentage as a function of time with varying intercepts across children. Fit using lmer() and interpret the coefficient for time.


```{r M1}

```

(b) Extend the model in (a) to include child-level predictors (that is, group-level predictors) for treatment and age at baseline. Fit using lmer() and interpret the coefficients on time, treatment, and age at baseline.

```{r group level predictor}

```

(c) Investigate the change in partial pooling from (a) to (b) both graphically and numerically.

```{r}

```

